---
title: "Remote sensing of Mountain pine beetle outbreaks; Darkwoods Conservation Area"
author: "SMurphy"
date: "2020-11-02"
output:
  github_document:
    toc: TRUE
    toc_depth: 5
    df_print: tibble
#always_allow_html: yes
---

```{r setup, include=FALSE, message = FALSE, warning=FALSE, error=FALSE}
options(htmltools.dir.version = FALSE, htmltools.preserve.raw = FALSE)
library(readxl)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(psych)
library(useful)
library(caret)
library(tibble)
library(DescTools)
library(animation)
library(ModelMetrics)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, 
  error=FALSE, comment = NA,tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## 1. Model Designs

### 1.1 Import Data

Set seeds to `123`. Import excel data representing training data of red-stage beetle attacks from candidate spectral indices sample in-field in forest health surveys.

```{r, message=FALSE, warning=FALSE}
set.seed(123) 
darkwoods_beetle_plots_data <- read_excel("2.1.darkwoods_beetle_ground_plots.xlsx")
print(darkwoods_beetle_plots_data)
```

### 1.2 Model Slope

```{r, message=FALSE, warning=FALSE}
beetle_ndmi <- lm(pi_mpb_killed_pc ~ ndmi, data = darkwoods_beetle_plots_data)
beetle_ndmi_residuals <- resid(beetle_ndmi)
summary(beetle_ndmi)

plot(pi_mpb_killed_pc ~ ndmi, data = darkwoods_beetle_plots_data, 
     main=NULL, 
     ylab = "Percent of basal area of pine trees killed by MPB", xlab = "NDMI", col="blue") +
  abline(lm(pi_mpb_killed_pc ~ ndmi, data = darkwoods_beetle_plots_data), col = "red")
```

### 1.3 Model Training

Splitting data 70:30 for training samples based on outcome variable: `pi_mpb_killed_pc`.

```{r, message=FALSE, warning=FALSE}
beetle_training.samples <- createDataPartition(darkwoods_beetle_plots_data$pi_mpb_killed, p=0.70, list = FALSE)
beetle_train.data <- darkwoods_beetle_plots_data[beetle_training.samples, ]
beetle_test.data <- darkwoods_beetle_plots_data[-beetle_training.samples, ]
```

Training regimes set with `time-slice` grid and 10K-fold cross validation with 10 repeats using `repeatedcv` function.

```{r, message=FALSE, warning=FALSE}
model_training_time_series <- trainControl(method = "timeslice",
                                           initialWindow = 36,
                                           horizon = 12,
                                           fixedWindow = TRUE)

model_training_10kfold <- trainControl(method = "repeatedcv", 
                                       number = 10, repeats=10)
                                       # animation of 10-kfold method:
animation::cv.ani(k=10)
```

### 1.4 Model Tuning

Models are fitted with the `svmLinear` support vector machine linear kernel and tuned with 3-step tuning grid and maximum of 20 hyperparameters. Data preprocessing included using `center` and `scale` functions.

```{r, message=FALSE, warning=FALSE}
#model 1 - NDMI - model specification
svm_ndmi_linear <- train(pi_mpb_killed ~ ndmi,
                         data = beetle_train.data,
                         method = "svmLinear",
                         trControl = model_training_10kfold,
                         preProcess = c("center","scale"),
                         tuneLength=10)

svm_ndmi_radial <- train(pi_mpb_killed ~ ndmi,
                         data = beetle_train.data,
                         method = "svmRadial",
                         trControl = model_training_10kfold, 
                         preProcess = c("center","scale"), 
                         tuneLength=10)

rf_ndmi_1000trees = train(pi_mpb_killed ~ ndmi,
                          data = beetle_train.data, 
                          method="rf", ntree=1000, 
                          metric="RMSE",
                          trControl=model_training_10kfold,
                          importance = TRUE)

```

### 1.5 Model Validation

```{r, message=FALSE, warning=FALSE}
#model 4 - results
beetle_ndmi_pred_train <- predict(svm_ndmi_linear, data = beetle_train.data)
beetle_ndmi_pred_train_mae <- mae(beetle_ndmi_pred_train, beetle_train.data$pi_mpb_killed)
beetle_ndmi_pred_train_mae
beetle_ndmi_pred_train_mae_rel <- (beetle_ndmi_pred_train_mae/mean(beetle_train.data$pi_mpb_killed))*100
beetle_ndmi_pred_train_mae_rel
beetle_ndmi_pred_train_rmse <- rmse(beetle_ndmi_pred_train, beetle_train.data$pi_mpb_killed)
beetle_ndmi_pred_train_rmse
beetle_ndmi_pred_train_rmse_rel <- (beetle_ndmi_pred_train_rmse/mean(beetle_train.data$pi_mpb_killed))*100
beetle_ndmi_pred_train_rmse_rel
beetle_ndmi_pred_train_R2 <- R2(beetle_ndmi_pred_train, beetle_train.data$pi_mpb_killed)
beetle_ndmi_pred_train_R2
TheilU(beetle_train.data$pi_mpb_killed, beetle_ndmi_pred_train, type = 2)
beetle_ndmi_pred_train_Ubias <- ((beetle_ndmi_pred_train_mae)*20)/((beetle_ndmi_pred_train_mae)^2)
beetle_ndmi_pred_train_Ubias

beetle_ndmi_pred_test <- predict(svm_ndmi_linear, data = darkwoods_beetle_plots_data)
beetle_ndmi_pred_test_rmse <- rmse(beetle_ndmi_pred_test, darkwoods_beetle_plots_data$pi_mpb_killed)
beetle_ndmi_pred_test_rmse/beetle_ndmi_pred_train_rmse
```

## 2. Full Model List

```{r,message=FALSE, warning=FALSE}

#model 2 - TAS-WET - model specification
svm_taswet_linear <- train(pi_mpb_killed ~ taswet,
                         data = beetle_train.data,
                         method = "svmLinear",
                         trControl = model_training_10kfold,
                         preProcess = c("center","scale"),
                         tuneGrid = expand.grid(C=seq(0,3, length = 20)))

#model 3 - TAS-GRE - model specification
svm_tasgreen_linear <- train(pi_mpb_killed ~ tasgre,
                         data = beetle_train.data,
                         method = "svmLinear",
                         trControl = model_training_10kfold,
                         preProcess = c("center","scale"),
                         tuneGrid = expand.grid(C=seq(0,3, length = 20)))

#model 4 - TAS-BRI - model specification
svm_tasbright_linear <- train(pi_mpb_killed ~ tasbri,
                         data = beetle_train.data,
                         method = "svmLinear",
                         trControl = model_training_10kfold,
                         preProcess = c("center","scale"),
                         tuneGrid = expand.grid(C=seq(0,3, length = 20)))

#model 1 results
svm_ndmi_linear
svm_ndmi_linear$finalModel
trellis.par.set(caretTheme())
densityplot(svm_ndmi_linear)

# model 2 results
svm_taswet_linear$finalModel
trellis.par.set(caretTheme())
densityplot(svm_taswet_linear)

# model 3 results
svm_tasgreen_linear
svm_tasgreen_linear$finalModel
trellis.par.set(caretTheme())
densityplot(svm_tasgreen_linear)

# model 4 results
svm_tasbright_linear
svm_tasbright_linear$finalModel
trellis.par.set(caretTheme())
densityplot(svm_tasbright_linear)
```

## 3. Appendix: Seperability Analysis of Spectral Bands & Indices

```{r, message=FALSE, warning=FALSE}
darkwoods_beetle_spectrals <- read_excel("2.2.darkwoods_beetle_spectral_sampling.xlsx")
describe(darkwoods_beetle_spectrals$NDMI_c)
mean(darkwoods_beetle_spectrals$NDMI_c)
SD(darkwoods_beetle_spectrals$NDMI_c)
median(darkwoods_beetle_spectrals$NDMI_c)
MeanSE(darkwoods_beetle_spectrals$NDMI_c)
shapiro.test(darkwoods_beetle_spectrals$NDMI_c)

mean(darkwoods_beetle_spectrals$TASWET)
SD(darkwoods_beetle_spectrals$TASWET)
median(darkwoods_beetle_spectrals$TASWET)
MeanSE(darkwoods_beetle_spectrals$TASWET)
shapiro.test(darkwoods_beetle_spectrals$TASWET)

mean(darkwoods_beetle_spectrals$TASBRI)
SD(darkwoods_beetle_spectrals$TASBRI)
median(darkwoods_beetle_spectrals$TASBRI)
MeanSE(darkwoods_beetle_spectrals$TASBRI)
shapiro.test(darkwoods_beetle_spectrals$TASBRI)

mean(darkwoods_beetle_spectrals$TASGRE)
SD(darkwoods_beetle_spectrals$TASGRE)
median(darkwoods_beetle_spectrals$TASGRE)
MeanSE(darkwoods_beetle_spectrals$TASGRE)
shapiro.test(darkwoods_beetle_spectrals$TASGRE)

mean(darkwoods_beetle_spectrals$B5NIR)
SD(darkwoods_beetle_spectrals$B5NIR)
median(darkwoods_beetle_spectrals$B5NIR)
MeanSE(darkwoods_beetle_spectrals$B5NIR)
shapiro.test(darkwoods_beetle_spectrals$B5NIR)

mean(darkwoods_beetle_spectrals$B1Aerosol)
SD(darkwoods_beetle_spectrals$B1Aerosol)
median(darkwoods_beetle_spectrals$B1Aerosol)
MeanSE(darkwoods_beetle_spectrals$B1Aerosol)
shapiro.test(darkwoods_beetle_spectrals$B1Aerosol)

mean(darkwoods_beetle_spectrals$B2Blue)
SD(darkwoods_beetle_spectrals$B2Blue)
median(darkwoods_beetle_spectrals$B2Blue)
MeanSE(darkwoods_beetle_spectrals$B2Blue)
shapiro.test(darkwoods_beetle_spectrals$B2Blue)

mean(darkwoods_beetle_spectrals$B7SW2)
SD(darkwoods_beetle_spectrals$B7SW2)
median(darkwoods_beetle_spectrals$B7SW2)
MeanSE(darkwoods_beetle_spectrals$B7SW2)
shapiro.test(darkwoods_beetle_spectrals$B7SW2)

mean(darkwoods_beetle_spectrals$B4Red)
SD(darkwoods_beetle_spectrals$B4Red)
median(darkwoods_beetle_spectrals$B4Red)
MeanSE(darkwoods_beetle_spectrals$B4Red)
shapiro.test(darkwoods_beetle_spectrals$B4Red)

mean(darkwoods_beetle_spectrals$B3Gree)
SD(darkwoods_beetle_spectrals$B3Gree)
median(darkwoods_beetle_spectrals$B3Gree)
MeanSE(darkwoods_beetle_spectrals$B3Gree)
shapiro.test(darkwoods_beetle_spectrals$B3Gree)

wilcox.test(B1Aerosol ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B2Blue ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B3Gree ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B4Red ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B5NIR ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B6SW1 ~ RA_NonRA, data = darkwoods_beetle_spectrals)
wilcox.test(B7SW2 ~ RA_NonRA, data = darkwoods_beetle_spectrals)
```


